---
title: 'Reproducible Research: Peer Assessment 1'
output: html_document
keep_md: yes
---

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data
``` {r activity,echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
# import the needful
library(dplyr)
library(tidyverse)
library(ggplot2)

# unzip and read in dataset, activity.
if (!file.exists("activity.csv")){
unzipped_file <- unzip("activity.zip", exdir = ".")}
activity <- read_csv("activity.csv")

# preview dataset
nrow(activity)
ncol(activity)
```

The activity dataset has been loaded. There are **17568** observations and **3** variables.

## What is mean total number of steps taken per day?
``` {r mean_gg, echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
# Calculate the total number of steps taken each day
total_steps <- aggregate(steps ~ date, data = activity, sum)

# A histogram to show
mean_gg <- ggplot(data = total_steps, aes(x = steps))+geom_histogram(binwidth = 3966.534)
mean_tot <- mean(total_steps$steps)
median_tot <- median(total_steps$steps)
mean_gg
```




The mean total number of steps taken each day is **10766** and the median is **10765**. We can see from the histogram that the distribution is not completely symmetrical, a bit skewed to the left of the distribution as also reflected in the mean being slightly higher than the median.There are also outliers in both ends of the distribution. Also, it is unimodal. 

## What is the average daily activity pattern?
``` {r timeseries, results='asis', comment=NA, warning=FALSE, message=FALSE}
total <- aggregate(steps ~ interval, data = activity, sum)
timeseries <- ggplot(data = total, aes(x = interval, y = steps))+geom_line()
timeseries
```
``` {r max, echo= TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
get_max <- filter(total, steps == max(total$steps))
get_step <- get_max[1,1]
get_value <- get_max[1,2]
```

On average, across all the days in the activity dataset, the 5-minute interval with the steps is **835** with a value of **10927**.

## Imputing missing values
``` {r missing, echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
miss <- is.na(activity)
```

There are **2304** number of missing values in the activity dataset.
``` {r avg, echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
# create a function, fun, that computes the mean number of steps for a day.
fun <- function(steps){
        s <- sum(steps)
        l <- length(steps)
        m <- s / l
}
avg <- aggregate(steps ~ date, data = activity, fun)
m_avg <- mean(avg$steps)

# replace na values with the computed mean
act <- replace_na(activity$steps, m_avg)

# create new dataset, activity_2
activity_2 <- activity
activity_2$steps <- act
```
We replace the missing values with the mean number of steps per day.
``` {r mean_g2, echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
# Calculate the total number of steps taken each day for our new dataset, activity_2
total_steps_2 <- aggregate(steps ~ date, data = activity_2, sum)

# A histogram to show
mean_g2 <- ggplot(data = total_steps_2, aes(x = steps))+geom_histogram(binwidth = 3523.595)
mean_tot2 <- mean(total_steps_2$steps)
median_tot2 <- median(total_steps_2$steps)
mean_g2
```


We have a more symmetric distribution as both the mean and median have the same values.

## Are there differences in activity patterns between weekdays and weekends?
``` {r weekdays, echo = TRUE, results='asis', comment=NA, warning=FALSE, message=FALSE}
# Change date format to define "weekends" and "weekdays" levels
activity_2$date <- weekdays(activity_2$date)

# Create a function, sec, that groups days into the defined levels
sec <- function(date){
if (date %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")){
        weektype <- "weekdays"
}
        else{
                weektype <- "weekends"
        }
}

# Create a column, weektype
activity_2 <- mutate(activity_2, weektype = "")
activity_2$weektype <- factor(sapply(activity_2$date, sec))

# Group our dataset by the weektype
total_2 <- aggregate(steps ~ interval + weektype, data = activity_2, fun)
weekdays <- ggplot(data = total_2, aes(x = interval, y = steps))+geom_line()+facet_wrap(~weektype, dir = "v", shrink = F)
weekdays
```


## Observations
+ On weekdays, the average number of steps for every 5 minute interval before the 1000 minute time interval is higher than the average number of steps on weekends.
+ On weekdays, the average number of steps for every 5 minute interval before the 1000 minute time interval is lower than the average number of steps on weekends.
+ Weekdays have a peak of about 205 average number of steps which is a lot higher than the peak average number of steps on weekends, 155.
